Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-21 ind_files/04_field_eqns.tex
Minor rephrasing

So far, we have dealt with populations in which individuals come in countably many different kinds. While developing these models, we have been on mathematically solid ground that is well understood by statistical physicists and mathematicians. However, things become more complicated when we deal with `quantitative' traits. Traits like body size, body weight, or beak length, often take on uncountably many values (say, all values in the interval $[0,1]$, for example). In this case, we cannot describe the population using a vector as we did before, but instead require a function. More precisely, if the set of all possible trait values is $\mathcal{T}$, we will characterize the population using a special kind of function $\phi^{(t)}$ such that the quantity $\int_{A}\phi^{(t)}(x)dx$ gives us the number of individuals that are in any `nice' region $A \subset \mathcal{T}$ of the possible trait space\footnote{The mathematically informed reader may notice that this sounds like I am trying to dance around the word `measure'. Indeed, we are really looking to construct branching processes that take values in some nice space of measures that can be endowed with sufficient mathematical structure for notions like convergence and integration to make sense. All the Dirac deltas that will turn up shortly are `properly' viewed as measures, and integrals with Dirac deltas in the integrand are to be interpreted as integration with respect to the Dirac measure. If one tries to be careful about these things, they will quickly find themselves drowning in a quagmire of mathematical formalism.
% For example, one can't really justify using informal tools like the functional derivative, which is ill-defined and `properly' interpreted as a Fr\'echet derivative in a suitable normed function space (making the use in conjunction with Dirac measures/distributions somewhat dubious without further elaboration on the space and norm in question). The existence of a probability density function for this process is also \emph{a priori} somewhat questionable and requires detailed arguments using tools like the Radon-Nikodym theorem. Here, we adopt the physicist's solution of ignoring all such potential difficulties and assuming a clever mathematician can make sure everything works out using arcane tools such as ``functional analysis" and  ``distribution theory".
If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$.

>>>>

So far, we have dealt with populations in which individuals come in countably many different kinds. While developing these models, we have been on mathematically solid ground that is well understood by statistical physicists and mathematicians. However, things become more complicated when we deal with `quantitative' traits. Traits like body size, body weight, or beak length, often take on uncountably many values (say, all values in the interval $[0,1]$, for example). In this case, we cannot describe the population using a vector as we did before, but instead require a function. More precisely, if the set of all possible trait values is $\mathcal{T}$, we will characterize the population using a special kind of function $\phi^{(t)}$ such that the quantity $\int_{A}\phi^{(t)}(x)dx$ gives us the number of individuals that are in any `nice' region $A \subset \mathcal{T}$ of the possible trait space\footnote{The mathematically informed reader may notice that this sounds like I am trying to dance around the word `measure'. Indeed, we are really looking to construct branching processes that take values in some nice space of measures that can be endowed with sufficient mathematical structure for notions like convergence and integration to make sense. All the Dirac deltas that will turn up shortly are `properly' viewed as measures, integrals with Dirac deltas in the integrand are `properly' interpreted as integration with respect to the Dirac measure, and the functional derivative is to be `properly' interpreted as a Gateaux or  Fr\'echet derivative assuming that the relevant function space has enough structure for these notions to make sense. If one tries to be careful about these things, they will quickly find themselves drowning in a quagmire of mathematical formalism.
% For example, one can't really justify using informal tools like the functional derivative, which is ill-defined and `properly' interpreted as a Fr\'echet derivative in a suitable normed function space (making the use in conjunction with Dirac measures/distributions somewhat dubious without further elaboration on the space and norm in question). The existence of a probability density function for this process is also \emph{a priori} somewhat questionable and requires detailed arguments using tools like the Radon-Nikodym theorem. Here, we adopt the physicist's solution of ignoring all such potential difficulties and assuming a clever mathematician can make sure everything works out using arcane tools such as ``functional analysis" and  ``distribution theory".
If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.} at time $t$. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$.
